Modal type theory based on the intuitionistic epistemic logic

### 2018-01-24 article.tex
fixes and erases

\begin{abstract}
Modal intuitionistic epistemic logic IEL$^{-}$ was proposed by S.Artemov and T. Protopopescu as the formal
foundation for the intuitionistic theory of knowledge. We construct a modal simply typed lambda-calculus
which is Curry-Howard isomorphic to IEL$^{-}$ as formal theory of calculations with applicative functors in
functional programming languages like Haskell or Idris. We prove that this typed lambda-calculus has the
strong normalization and Church-Rosser properties.
\end{abstract}

@@

@@

      \begin{prooftree}
      \AxiomC{ $\Gamma \vdash M : A$ }
      \AxiomC{ $\Gamma \vdash N : B$ }
      \RightLabel{$\times_i$}
      \BinaryInfC{$\Gamma \vdash \langle x, y \rangle : A \times B$}
      \end{prooftree}

@@

\begin{itemize}
\item $(\lambda x. {\bf pure \: } x) : A \to \textbf{K}A$;
\item $\lambda f. \lambda x. {\bf let \: pure} \: \langle g, y \rangle = \langle f, x \rangle {\: \bf in \:} g y : {\bf K}(A \to B) \to {\bf K} A \to {\bf K} B$
\item $\lambda f. \lambda x. {\bf let \:pure} \: \langle g, y \rangle = \langle {\bf pure \:} f, x \rangle {\: \bf in \: } g y : (A \to B) \to {\bf K} A \to {\bf K} B$
\end{itemize}

@@

  6) $\lambda x. f x \rightarrow_{\eta} f$;

  7) $\langle \pi_1 P, \pi_2 P \rangle \rightarrow_{\eta} P$;

  8) ${\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N \rightarrow_{\eta} {\bf pure \:} N$;

@@

\begin{lemma} Generation lemma.

  ii) Let $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K}B$, there are some $A_1,\dots,A_n \in \mathbb{T}_{{\bf K}}$,
  such that $\Gamma \vdash \vec{M} : {\bf K} \vec{A}$ and $\vec{x} : \vec{A} \vdash N : B$.
\end{lemma}

  Induction on $\Gamma \vdash {\bf pure \:} M : {\bf K}A$ and $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{N} {\: \bf in \:} N : {\bf K}B$ correspondently.
\end{proof}

\begin{lemma} Weakening.

  Let $\Gamma \vdash M : A$ and $\Gamma \subseteq \Delta$, then $\Delta \vdash M : A$.

  1) Let $\Gamma, x : A \vdash x : A$ and $\Gamma \subseteq \Delta$, then $\Delta, x : A \vdash x : A$ trivially.

  2) Let $\Gamma \vdash {\bf pure \:} M : {\bf K}A$. Then $\Gamma \vdash M : A$ by generation and $\Delta \vdash M : A$ by assumption. So $\Delta \vdash {\bf pure \:} M : {\bf K} A$ by ${\bf K}_I$.

  3) Let $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K}B$ and $\Gamma \subseteq \Delta$. Then $\Gamma \vdash \vec{M} : {\bf K}\vec{A}$ and $\vec{x} : \vec{A} \vdash N : B$.

  By assumption $\Delta \vdash \vec{M} : {\bf K}\vec{A}$. So $\Delta \vdash {\bf let \: pure \:} \vec{x} = \vec{N} {\: \bf in \:} N : {\bf K}B$ by $\text{let}_{{\bf K}}$.
\end{proof}

@@

\begin{lemma} Substituition lemma.

    i) Let $\Gamma, x : A \vdash M : B$ and $\Gamma \vdash N : A$, then $\Gamma \vdash M [x := N] : B$.

    ii) Let $\Gamma \vdash M : A$, then $\Gamma [B := C] \vdash M : (A [B := C])$.

i) For term substitution:

1) Let $\Gamma, x : A \vdash x : A$ and $\Gamma \vdash N : A$, but $x [x := N] = N$, so $\Gamma \vdash N : A$.

2) Let $\Gamma, x : A \vdash {\bf pure \:} M : {\bf K}B$ and $\Gamma \vdash N : A$.

@@

3) Let $\Gamma, y : A \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K} B$ and $\Gamma \vdash N : A$.

By generation, $\Gamma, y : A \vdash \vec{M} : {\bf K} \vec{A}$ and $\vec{x} : \vec{A} \vdash N : B$.

By hypothesis, $\Gamma \vdash \vec{M}[x := N] : {\bf K} \vec{A}$.

Hence $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M}[x := N] {\: \bf in \:} N : {\bf K} B$.

ii) For type substitution

1) Let $\Gamma, x : A \vdash x : A$, so $\Gamma [A := C], x : (A[A := C]) \vdash x : (A[A := C])$, or

$\Gamma [A := C], x : C \vdash x : C$.

2) Let $\Gamma \vdash {\bf pure \:} M : {\bf K}A$. By generation $\Gamma \vdash M : A$ and by assumption $\Gamma [B := C] \vdash M : A[B := C]$.

By $K_I$ $\Gamma \vdash {\bf pure \:L} M : {\bf K}(A[B := C])$.

3) $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K}B$. By generation $\Gamma \vdash \vec{M} : {\bf K}\vec{A}$ and $\vec{x} : \vec{A} \vdash N : B$.

By assumption $\Gamma [B_1 := C] \vdash \vec{M} : {\bf }{K}\vec{A}[B_1 := C]$ and $\vec{x} : \vec{A}[B_1 := C] \vdash N : B[B_1 := C]$.

So by $\text{let}_{{\bf K}}$, $\Gamma [B_1 := C] \vdash {\bf let \: pure \:} \vec{x} = \vec{M} {\: \bf in \:} N : {\bf K}(B[B_1 := C])$.

@@

3) Let $\Gamma \vdash {\bf let \: pure \:} \vec{x} = {\bf pure \:} \vec{M} {\: \bf in \:} N : {\bf K}B$.

By generation $\Gamma \vdash {\bf pure \:} \vec{M} : {\bf K}\vec{A}$ and $\vec{x} : \vec{A} \vdash N : B$.
\\ Moreover, $\Gamma \vdash \vec{M} : \vec{A}$. By weakening and substitution lemma $\Gamma \vdash N [\vec{x} = \vec{M}] : B$.
\\ By ${\bf K}_I$, $\Gamma \vdash {\bf pure \:} N [\vec{x} := \vec{M}] : {\bf K}B$.

4) Let $\vdash {\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N : {\bf K}A$

By generation $\vdash N : A$. \\ So $\vdash {\bf pure \:} N : {\bf K}A$ by ${\bf K}_I$.

5) Let $\Gamma \vdash {\bf pure \:} M : A$ and $M \twoheadrightarrow_{\beta \eta} N$.

By generation $\Gamma \vdash M : A$ and $\Gamma \vdash N : A$ by assumption.
\\ So $\Gamma \vdash {\bf pure \:} N : {\bf K}A$.
\end{proof}

@@

We will consider term model for simply typed lambda calculus $\times$ and $\to$ standardly described in \cite{LambekScott} \cite{Abramsky}.

@@

  i) ${\bf K}(g \circ f) = {\bf K}(g) \circ {\bf K}(f)$;

  ii) ${\bf K}(id_A) = id_{{\bf K}A}$.
\end{lemma}

@@

  \bibitem{LambekScott}

  \bibitem{Abramsky}

  \bibitem{Gabbay}

>>>>

\begin{abstract}
Modal intuitionistic epistemic logic IEL$^{-}$ was proposed by S.Artemov and T. Protopopescu as the formal
foundation for the intuitionistic theory of knowledge. We construct a modal simply typed lambda-calculus
which is Curry-Howard isomorphic to IEL$^{-}$ as formal theory of calculations with applicative functors in
functional programming languages like Haskell or Idris.
\end{abstract}

@@

  The motivation for using an applicative functor is quite natural. Applicative functor allows to generalize the action of a functor for functions with arbitrary number of arguments, for instance:
  \begin{lstlisting}[language=Haskell]
  liftA2 :: Applicative f => (a -> b -> c) -> f a -> f b -> f c
  liftA2 f x y = pure f <*> x <*> y
  \end{lstlisting}

@@

      \begin{prooftree}
      \AxiomC{ $\Gamma \vdash M : A$ }
      \AxiomC{ $\Gamma \vdash N : B$ }
      \RightLabel{$\times_i$}
      \BinaryInfC{$\Gamma \vdash \langle M, N \rangle : A \times B$}
      \end{prooftree}

@@

\begin{itemize}
\item $(\lambda x. {\bf pure \: } x) : A \to \textbf{K}A$;
\item $\lambda f. \lambda x. {\bf let \: pure} \: g, y  = f, x {\: \bf in \:} g y : {\bf K}(A \to B) \to {\bf K} A \to {\bf K} B$
\item $\lambda f. \lambda x. {\bf let \:pure} \: g, y = {\bf pure \:} f, x  {\: \bf in \: } g y : (A \to B) \to {\bf K} A \to {\bf K} B$
\end{itemize}

@@

  6) ${\bf let \: pure \:} \underline{\quad} = \underline{\quad} {\: \bf in \:} N \rightarrow_{\beta} {\bf pure \:} N$;

  7) $\lambda x. f x \rightarrow_{\eta} f$;

  8) $\langle \pi_1 P, \pi_2 P \rangle \rightarrow_{\eta} P$;

@@

@@

\begin{lemma} Basic lemmas for ${\bf K}_I$.

  ii) Let $\Gamma \vdash M : A$ and $\Gamma \subseteq \Delta$, then $\Delta \vdash M : A$;

  iii) Let $\Gamma, x : A \vdash {\bf pure \:} M : {\bf K}B$ and $\Gamma \vdash N : A$, then $\Gamma \vdash {\bf pure \:} M [x := N] : {\bf K}B$.

  iv) Let $\Gamma \vdash {\bf pure \:} M : {\bf K}A$, then $\Gamma [B := C] \vdash {\bf pure \:} M : {\bf K} (A [B := C])$.
\end{lemma}

  i) Induction on $\Gamma \vdash {\bf pure \:} M : {\bf K}A$ and $\Gamma \vdash {\bf let \: pure \:} \vec{x} = \vec{N} {\: \bf in \:} N : {\bf K}B$ correspondently.

  ii) Let $\Gamma \vdash {\bf pure \:} M : {\bf K}A$. Then $\Gamma \vdash M : A$ by generation and $\Delta \vdash M : A$ by assumption. So $\Delta \vdash {\bf pure \:} M : {\bf K} A$ by ${\bf K}_I$.

  iii) Let $\Gamma, x : A \vdash {\bf pure \:} M : {\bf K}B$ and $\Gamma \vdash N : A$.

@@

  iv) Let $\Gamma \vdash {\bf pure \:} M : {\bf K}A$. By generation $\Gamma \vdash M : A$ and by assumption $\Gamma [B := C] \vdash M : A[B := C]$.

  By $K_I$ $\Gamma \vdash {\bf pure \:L} M : {\bf K}(A[B := C])$.
\end{proof}

@@

@@

We will consider term model for simply typed lambda calculus $\times$ and $\to$ standardly described in \cite{LambekScott}.

@@

  i) $\text{fmap }(g \circ f) = \text{fmap }(g) \circ \text{fmap }(f)$;

  ii) $\text{fmap }(id_A) = id_{{\bf K}A}$.
\end{lemma}

@@

  \bibitem{LambekScott} Lambek, J. and Scott P.J. (1986) Introduction to Higher Order Categorical Logic, Cambridge Studies in Advanced Mathematics 7, Cambridge: Cambridge University Press.

  \bibitem{ElKelly} Samuel Eilenberg and Max Kelly, Closed categories. Proc. Conf. Categorical Algebra (La Jolla, Calif., 1965).
